Towards a framework for reasoning about stateful compiler IRs using coinductive reasoning.

### 2024-12-05 paper.tex
Update paper.tex
document my internship experience

I'm planning to write a paper about this work tentatively titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
in which we explore all of these different strategies, 
classify exactly which class of expressions they solve, 
and investigate how many real-world problems are already covered by these strategies.
The aim is to submit this to PLDI, which has its call for papers around November.

@@

Furthermore, I plan to continue this work on bitvector automation in
Lean during my internship at Amazon\footnote{Assuming I'll get
  permission to work away, of course}. They are particularly interested
in automatically reasoning about ARM assembly code, which closely aligns
with our goal of automating bitvector reasoning to then use for
verifying (low-level) MLIR dialects.

This internship will thus provide me with exactly the corpus of real-world bitvector verification problems that I need to evaluate the strategies we'll discuss in the 
aforementioned paper.

>>>>

This culminated in a paper titled: ``\emph{Taming the Bitvector Bestiary: A Characterization of Whitebox Bitvector Proof Strategies}'', 
which we submitted to PLDI.

@@

At Amazon, I wrote a scalable symbolic simulator for straight-line, memory-free Arm machine code, in the context of LNSym, 
a toolkit for the verification of cryptographic hash functions and block ciphers written in Arm assembly.
Before my internship, a prototype symbolic simulator existed, but it was limited in features and did not scale well.
My reimplemented version tracks more information of the simulated execution, 
and is capable of simulating more than 400 instructions of the SHA512 litmus test we benchmarked against before hitting Lean's default time-out limits.
This is about 10x more instructions than the prototype was able to simulate given the same time-out limits.

The prototype relied heavily on Lean's builtin simplifier. 
I replaced this with bespoke metacode, that was able to exploit the specific shape of the problem we were solving.
Firstly, I wrote preprocessing that allowed us to treat the operational semantic model, where operations are given semantics as functions that write to state, 
in an axiomatic-semantics way, by having theorems that characterizes operations in terms of reads from the resulting state.
Secondly, I wrote a bespoke data-structure and used it selectively reflects parts of the proof goal.
We explicitly decoupled this meta-code needed to scale symbolic simulation from the definition of the semantic model,
so that the model remains extensible even by people who are not experts in formal verification, which was an explicit design goal.

This project, like some of my other research, was also an experiment in the usability of Lean as a platform to verify software.
In particular, we generated very large proofs, and exposed, amongst others, a fundamental scalability problem in Lean:
The time Lean needs to assemble a proof after a tactic executes scales roughly by the length of the proof times the number of binders in the proof.
In the larger benchmarks, LNSym would generate proofs involving tens of thousands of binders,
resulting in the majority of time needed to check the proof was actually spent in this proof-assembling step.
We reported this to the Lean core team, who are actively working on a solution.
Furthermore, the semantic model of Arm machine code used relies heavily on bitvectors, thus, we also used the proof obligations resulting from the improved symbolic simulator to stress-test LeanSAT. I've reported these problems, and submitted patches for some, contributing to the greater vision of improving the usability of Lean.

Besides the concrete contribution of the code I wrote, I also documented my experience,
provifing an explanation of the various optimizations I tried, and which worked or didn't worked,
resulting in a clear roadmap for the LNSym team to iterate on.
